Turbulence and Particle Acceleration in Giant Radio Halos: the Origin of Seed Electrons

### 2016-04-21 paper.tex
Added files via upload

@@

% --- section: Analytical Results and discussion --- %

\section{Parameter space exploration with spherical models}
\label{sect:param_comp}
In this section we explore the robustness of our models to crucial
parameters. Figure~\ref{fig:param_comp} shows how assumptions about
turbulence and CRs impact the radio emission. We find that the level
of turbulence has a large impact on both radio spectra and profile,
while the parameter $\alpha_\rmn{tu}$ that determines the spatial
profile of turbulence mainly influence the radio
profiles. Interestingly, assumptions about CRs have less impact on
radio emission. This can be understood by looking at the scaling of
parameters to the reacceleration timescale, where $\tau_\rmn{D}
\propto X_\rmn{tu}^2$. The CRps approximately gain a factor $\Delta
C_\rmn{reacc.} \sim
\exp{\left[\frac{(2+\alpha)(1-\alpha)}{4}\frac{\tau_\rmn{cl}}{\tau_\rmn{D}}\right]}$
during reacceleration if injection and cooling are neglected. The CRes
gain a similar factor from reacceleration, however in the region
around $0.1\,R_{200}$ the Coulomb cooling timescale
($\tau_\rmn{coul}\approx0.01\,\rmn{Gyr}$) is much shorter than
$\tau_\rmn{D}\approx0.5\,\rmn{Gyr}$, hence $\Delta C_\rmn{reacc.}$
will be greatly reduced for the electrons. The spatial region is
important when we derive the radio spectrum, since in order to compare
to observations we integrate the CR profile out to $\sim
0.2\,R_{200}$. For a CRp spectral index $\alpha_\rmn{CR}\sim 2.5$,
$\Delta C_\rmn{reacc.} \sim
\exp{\left[1.7\frac{\tau_\rmn{cl}}{\tau_\rmn{D}}\right]} \sim 10$ for
$X_\rmn{tu}=0.2$ and $\alpha_\rmn{tu}=0.8$. This is confirmed by the
top panels in Fig.~\ref{fig:param_comp}. Increasing from
$X_\rmn{tu}=0.2$ to $X_\rmn{tu}=0.3$, gives us a $\Delta
C_\rmn{reacc.}  \sim 100$, which is roughly confirmed by same
figures. The factor 2-3 difference in the figure is due to the shorter
reacceleration timescale that increase the contribution from
reaccelerated CRes. Hence we conclude that the radio emission is much
more sensitive to the exponential dependence on turbulence than the
linjear dependence on injected CRs. \AP{Too complicated, perhaps cut
  or simplify?}

@@

% --- section: Simulation Results and discussion --- %

\section{Results of cosmological simulations}
\label{sec:results}

In this section we show that our three models individually reproduce
the observed Coma radio halo without violating gamma-ray bounds. The
level of emission is mainly driven by how efficient the CRs are
reaccelerated and for how long. After turbulent reacceleration, the
volume-weighted, relative CRp energy density and relative CRp number
density inside the RH for \Mflatturb (\Mstream), are found to be 2 (3)
\% and $2\times10^{-8}$ ($5\times10^{-8}$), respectively. As we will
see later, these densities are just of the right order of magnitude to
reproduce radio observations in the Coma cluster.

\AP{move to a scaling subsection?} The diffusion constant,
$D_\rmn{pp}\propto I_0/\rho c_s \propto
X_\rmn{tu}^2k_0\epsilon_\rmn{th}^{\alpha_\rmn{tu}-1}\sqrt{T}$, where
we adopt the same level of compressible turbulence, $X_\rmn{tu}=0.2$,
for all models. The reaccelerated CRs are exponentially dependent on
this parameter as long as $\tau_D > \tau_\rmn{cl}$, where
$\tau_\rmn{cl}$ represent the duration of reacceleration of the merger
which we assume to be 650~Myrs long. Note that a fixed reacceleration
timescale $\tau_D\sim 1/D_\rmn{pp}\sim 0.2-0.7\,\rmn{Gyr}$ is required
to explain the observations for each model and radial bin. This allows
us to interchange $\tau_\rmn{cl} \propto X_\rmn{tu}^2 \propto k_0$,
i.e. a longer reacceleration duration have to be compensated by
lowering the level of turbulence or increasing the injection scale of
turbulence to keep $\tau_D$ constant. We also exploit the large
uncertainty in the turbulent profile and adjust $\alpha_\rmn{tu}$ for
each model so that the CRs have the correct spatial profile to match
the radio emission. The shape of the spectrum for each of model is
more complicated and is determined both from reacceleration and as
well as the gains and losses during the formation of the cluster. It
depends only weakly on the specifics of the CR injection at the shocks
(where the Mach number distribution of the shocks is the most
important quantity). The normalization of the spectrum is mainly
determined from the maximum acceleration efficiency (together with the
uncertain parameters $X_\rmn{tu}$, $\tau_\rmn{cl}$, and $k_0$). Given
the freedom in the turbulens, it might not be surprising that we match
the radio profile in Coma. Although, matching the spectrum without
violating gamma-ray bounds for each of our three models is more
surprising and is not a consequence of parameter fitting. We believe
that it is ultimately a consequence of the adopted physics and hint
that reacceleration is a viable explanation for giant radio
halos. Below we discuss in more detail the different emission
components of Coma.

@@

\subsection{Radio spectrum}
In Fig.~\ref{fig:sync_spectrum} we show that our three models that
include Fermi-II reacceleration can individually reproduce the
convexly curved total radio spectrum found in the Coma cluster. Seed
CRs in \Mstream and \Mflatturb that do not experience turbulent
reacceleration have a power-law spectrum in disagreement with
observations. In order to match both the spatial and spectral profiles
in Coma, we adopt an acceleration efficiency for the strongest shocks
in our three models \Mprimary, \Mstream, and \Mflatturb to
$\zeta_{\rmn{e}} <0.003$, $\zeta_{\rmn{p}} < 0.1$, and
$\zeta_{\rmn{p}}<0.03$, respectively. Following the Mach number
($\mathcal{M}$)-dependence of the acceleration efficiency suggested in
\cite{pinzke13}, the efficiency in weak shocks ($\mathcal{M}\sim
2.5-3.5$) that dominates the CR distribution function, has an
acceleration efficiency for protons $\zeta_{\rmn{p}}\sim0.0001-0.01$,
and for electrons $\zeta_{\rmn{e}}\sim 0.001$. Interestingly, we find
that the radio luminosity from clusters in the OFF-state (DSA only)
and ON-state (DSA and reacceleration) differ by about a factor 10 in
all our three models. However, note that for \Mprimary, the primary
CRes that generate most of the radio emission from the cluster in the
OFF-state are dominated by only a small fraction of the CRes. These
electrons are injected very recently and have not had time to cool
yet. Hence we expect there to be a large variance in the OFF-state of
different simulated clusters.

While the radio surface brightness has an approximate linear
dependence on the acceleration efficiency, it has a strong non-linear
dependence on the level of compressible turbulence. Below a critical
level ($X_\rmn{thres,tu} = \frac{400\,\rmn{Myrs}}{\tau_\rmn{cl}}
X_\rmn{tu}\approx 0.12$ in our fiducial model), turbulent
reacceleration is negligible, while it rises sharply above this
threshold. For instance, for $X_\rmn{tu}\gtrsim0.12$, a factor two
change in $X_\rmn{tu}$ changes the radio surface brightness by $\sim
10-100$ (see also section~\ref{sect:param_comp}). In principle this
threshold behavior allows us to provide a strict lower bound on
turbulence in radio halos. An even more stringent bound on
$X_\rmn{tu}$ can be inferred from combining radio observations with
the gamma-ray observations where the predicted fluxes from both
\Mstream and \Mflatturb are close to the upper limits set by
Fermi-LAT. If $X_\rmn{tu}$ is smaller than in our adopted models
(where we assume $X_\rmn{tu}=0.2$), then the efficiency of DSA has to
be larger than $\zeta_\rmn{p}\sim 0.1$ for the secondary CRes to
reproduce the radio observations. However, since the turbulent
reacceleration acts on both the secondary CRes and the CRps, while
$\zeta_\rmn{p}$ only affects the CRps, these models would produce too
much gamma-rays. Hence we conclude that $X_\rmn{tu}\gtrsim0.2$ if all
other parameters are kept fixed. Although, we caution the reader to
take this limit too stringent since there are degeneracies between the
uncertain parameters $X_\rmn{tu}\propto\sqrt{\tau_\rmn{D}}\propto
\sqrt{k_0}$. We will explore this parameter space further in future
work.

@@

\subsection{Gamma-rays}
The gamma-ray emission from CRps that produce decaying neutral pions
could be substantial if the CRps are reaccelerated efficiently enough,
hence it is interesting to estimate this emission for our models and
compare to upper limits. We predict the gamma-ray emission from
\Mflatturb (\Mstream) with $F_\gamma(>500\,\rmn{MeV})=4\times10^{-10}
(5\times10^{-10}) \mathrm{ \,ph\, s}^{-1}\mathrm{cm}^{-2}$. The fluxes
from these models are slightly larger than in \cite{brunetti12}, where
the differences comes from our steeper CRp profiles in addition to
simulation based formalism we rely on that accounts for both Coulomb
and hadronic losses during the build up of the CR distribution in
contrast to the scaling relations adopted in their
paper. Interestingly the gamma-ray flux from both our scenarios are
just below recent Fermi-LAT limits derived from a gamma-ray profile
similar to \Mflatturb where
$F_\gamma(>500\,\rmn{MeV})<5.3\times10^{-10} \mathrm{ \,ph\,
  s}^{-1}\mathrm{cm}^{-2}$ \footnote{Fabio Zandanel, private
  communication. See also
  \citet{2014MNRAS.440..663Z,2014ApJ...787...18A}. and will be probed
  in the next few years by Fermi-LAT. } The spectral index of the CRp
distribution is relatively steep ($\alpha_{\rmn{p}}\sim2.6$) for the
CRp energies $E \gtrsim 10$~GeV that are relevant for the injection of
radio-emitting secondary CRes. The steep spectrum is ultimately a
consequence \AP{of the shock history of the simulated cluster, with a
  weak dependence on} our test particle model for Fermi-I acceleration
\citep{pinzke13}, where we steepen the spectral index to avoid
acceleration efficiencies above $\zeta_{\rmn{p}} \sim 10\%$.

>>>>

\subsection{Scaling relations for turbulent reacceleratio}
The diffusion constant, $D_\rmn{pp}\propto I_0/\rho c_s \propto
X_\rmn{tu}^2k_0\epsilon_\rmn{th}^{\alpha_\rmn{tu}-1}\sqrt{T}$, where
we adopt the same level of compressible turbulence, $X_\rmn{tu}=0.2$,
for all models. The reaccelerated CRs are exponentially dependent on
this parameter as long as $\tau_D > \tau_\rmn{cl}$, where
$\tau_\rmn{cl}$ represent the duration of reacceleration of the merger
which we assume to be 650~Myrs long. Note that a fixed reacceleration
timescale $\tau_D\sim 1/D_\rmn{pp}\sim 0.2-0.7\,\rmn{Gyr}$ is required
to explain the observations for each model and radial bin (see
Table~\ref{tab:timescales}). This allows us to interchange
$\tau_\rmn{cl} \propto X_\rmn{tu}^2 \propto k_0$, i.e. a longer
reacceleration duration have to be compensated by lowering the level
of turbulence or increasing the physical injection scale of turbulence
to keep $\tau_D$ constant.

@@

% --- section: Analytical Results and discussion --- %
\section{Parameter space exploration with spherical models}
\label{sect:param_comp}
in this paper we rely on several critical parameters decribing
relatively unknown non-thermal physics in the ICM. Here we try to
explore the robustness of our models to the most critical parameters
related to turbulence and CRs. Figure~\ref{fig:param_comp} shows their
impact on radio emission. We find that the level of turbulence
($X_\rmn{tu}$) has a large impact on both radio spectra and profile,
while the parameter $\alpha_\rmn{tu}$ that determines the spatial
profile of turbulence mainly influence the radio profiles. As
expected, the radio profiles flatten significantely with a flatter
turbulent profile. A higher amount of energy also flatten the radio
profile, where the reason beeing that the steep profiles of injected
CRs have less impact on the radio emission for an increasing
reacceleration efficiency. Interestingly, assumptions about CRs have
less impact on radio emission. This can be understood by looking at
the scaling of parameters to the reacceleration timescale, where
$\tau_\rmn{D} \propto X_\rmn{tu}^2$. The CRps approximately gain a
factor $\Delta C_\rmn{reacc.} \sim
\exp{\left[\frac{(2+\alpha)(1-\alpha)}{4}\frac{\tau_\rmn{cl}}{\tau_\rmn{D}}\right]}$
during reacceleration if injection and cooling are neglected. The CRes
gain a similar factor from reacceleration, however in the region
around $0.1\,R_{200}$ the Coulomb cooling timescale
($\tau_\rmn{coul}\approx0.01\,\rmn{Gyr}$) is much shorter than
$\tau_\rmn{D}\approx0.5\,\rmn{Gyr}$, hence $\Delta C_\rmn{reacc.}$
will be greatly reduced for the electrons. The spatial region is
important when we derive the radio spectrum, since in order to compare
to observations we integrate the CR profile out to $\sim
0.2\,R_{200}$. For a CRp spectral index $\alpha_\rmn{CR}\sim 2.5$,
$\Delta C_\rmn{reacc.} \sim
\exp{\left[1.7\frac{\tau_\rmn{cl}}{\tau_\rmn{D}}\right]} \sim 10$ for
$X_\rmn{tu}=0.2$ and $\alpha_\rmn{tu}=0.8$. This is confirmed by the
top panels in Fig.~\ref{fig:param_comp}. Increasing the level of
turbulence to $X_\rmn{tu}=0.3$, gives us a $\Delta C_\rmn{reacc.}
\sim 100$, which is roughly confirmed by same figures. The factor 2-3
difference in the figure is due to the shorter reacceleration
timescale that increase the contribution from reaccelerated CRes that
would othervise cool away on a much shorter timescale. Hence we
conclude that the radio emission is much more sensitive to the
exponential dependence on compressible turbulence than the linjear
dependence on injected CRs. \AP{Too complicated? Perhaps cut or
  simplify?}

We can also find a lower critical level of turbulence where
reacceleration is negliable, while it rises sharply above this
threshold. We define the threshold by $\Delta C_\rmn{reacc.}  \lesssim
2$, which gives us an $X_\rmn{thres,tu} = \left(\frac{\log(2)
  0.04}{1.7}\frac{\tau_\rmn{D}}{\tau_\rmn{cl}}\right)^{0.5} \approx 0.08$ for
$\tau_\rmn{D} = 0.5\,\rmn{Myr}$. This threshold is also seen in the
top panels of figure~\ref{fig:param_comp}. For instance, for
$X_\rmn{tu}\gtrsim0.08$, a factor two change in $X_\rmn{tu}$ changes
the radio surface brightness by a factor $\sim 10-100$. In principle
this threshold behavior allows us to provide a strict lower bound on
turbulence in radio halos. An even more stringent bound on
$X_\rmn{tu}$ can be inferred from combining radio observations with
the gamma-ray observations (see section~\ref{sect:radio_spec}).

@@

% --- section: Simulation Results and discussion --- %
\section{Results of cosmological simulations}
\label{sec:results}

In this section we show that our three models individually reproduce
the observed Coma radio halo without violating gamma-ray bounds. The
level of emission is mainly driven by how efficient the CRs are
reaccelerated and for how long. After turbulent reacceleration, the
volume-weighted, relative CRp energy density and relative CRp number
density inside the RH for \Mflatturb (\Mstream), are found to be 2 (3)
\% and $2\times10^{-8}$ ($5\times10^{-8}$), respectively. As we will
see later, these densities are just of the right order of magnitude to
reproduce radio observations in the Coma cluster. We exploit the large
uncertainty in the turbulent profile and adjust $\alpha_\rmn{tu}$ for
each model so that the CRs have the correct spatial profile to match
the radio emission. The shape of the spectrum for each of model is
more complicated and is determined both from reacceleration and as
well as the gains and losses during the formation of the cluster. It
depends only weakly on the specifics of the CR injection at the shocks
(where the Mach number distribution of the shocks is the most
important quantity). The normalization of the spectrum is mainly
determined from the maximum acceleration efficiency (together with the
uncertain parameters $X_\rmn{tu}$, $\tau_\rmn{cl}$, and $k_0$). Given
the freedom in the turbulens, it might not be surprising that we match
the radio profile in Coma. Although, matching the spectrum without
violating gamma-ray bounds for each of our three models is more
surprising and is not a consequence of parameter fitting. We believe
that it is ultimately a consequence of the adopted physics and hint
that reacceleration is a viable explanation for giant radio
halos. Below we discuss in more detail the different emission
components of Coma.

@@

\subsection{Radio spectrum}
\label{sect:radio_spec}
In Fig.~\ref{fig:sync_spectrum} we show that our three models that
include Fermi-II reacceleration can individually reproduce the
convexly curved total radio spectrum found in the Coma cluster. Seed
CRs in \Mstream and \Mflatturb that do not experience turbulent
reacceleration have a power-law spectrum in disagreement with
observations. In order to match both the spatial and spectral profiles
in Coma, we adopt an acceleration efficiency for the strongest shocks
in our three models \Mprimary, \Mstream, and \Mflatturb to
$\zeta_{\rmn{e}} <0.003$, $\zeta_{\rmn{p}} < 0.1$, and
$\zeta_{\rmn{p}}<0.03$, respectively. Following the Mach number
($\mathcal{M}$)-dependence of the acceleration efficiency suggested in
\cite{pinzke13}, the efficiency in weak shocks ($\mathcal{M}\sim
2.5-3.5$) that dominates the CR distribution function, has an
acceleration efficiency for protons $\zeta_{\rmn{p}}\sim0.0001-0.01$,
and for electrons $\zeta_{\rmn{e}}\sim 0.001$. Interestingly, we find
that the radio luminosity from clusters in the OFF-state (DSA only)
and ON-state (DSA and reacceleration) differ by about a factor 10 in
all our three models. However, note that for \Mprimary, the primary
CRes that generate most of the radio emission from the cluster in the
OFF-state are dominated by only a small fraction of the CRes. These
electrons are injected very recently and have not had time to cool
yet. Hence we expect there to be a large variance in the OFF-state of
different simulated clusters. As mentioned in
section~\ref{sect:param_comp}, combining radio observations with
gamma-ray limits allows us to put a lower limit to $X_\rmn{tu}$.  If
$X_\rmn{tu}$ is smaller than in our adopted models (where we assume
$X_\rmn{tu}=0.2$), then the efficiency of DSA has to be larger than
$\zeta_\rmn{p}\sim 0.1$ for the secondary CRes to reproduce the radio
observations. However, since the turbulent reacceleration acts on both
the secondary CRes and the CRps, while $\zeta_\rmn{p}$ only affects
the CRps, \Mstream and \Mflatturb would produce too much
gamma-rays. Hence we conclude that $X_\rmn{tu}\gtrsim0.2$ if all other
parameters are kept fixed. Although, we caution the reader to take
this limit too stringent because of the uncertainty in $k_0$ and
$\tau_\rmn{cl}$ that impact $X_\rmn{tu}$ for a fixed
$\tau_\rmn{D}$. This parameter space needs to be explored further in
future work in order to put more stringent limits on the level of
turbulence in clusters using radio and gamma-ray observations in
combination with turbulent reaccelerated CRs.

@@

\subsection{Gamma-rays}
The gamma-ray emission from CRps that produce decaying neutral pions
could be substantial if the CRps are reaccelerated efficiently enough,
hence it is interesting to estimate this emission for our models and
compare to upper limits. We predict the gamma-ray emission from
\Mflatturb (\Mstream) with $F_\gamma(>500\,\rmn{MeV})=4\times10^{-10}
(5\times10^{-10}) \mathrm{ \,ph\, s}^{-1}\mathrm{cm}^{-2}$. The fluxes
from these models are slightly larger than in \cite{brunetti12}, where
the differences comes from our steeper CRp profiles in addition to
simulation based formalism we rely on that accounts for both Coulomb
and hadronic losses during the build up of the CR distribution in
contrast to the scaling relations adopted in their
paper. Interestingly the gamma-ray flux from both our scenarios are
just below recent Fermi-LAT limits derived from a gamma-ray profile
similar to \Mflatturb where
$F_\gamma(>500\,\rmn{MeV})<5.3\times10^{-10} \mathrm{ \,ph\,
  s}^{-1}\mathrm{cm}^{-2}$ \footnote{Fabio Zandanel, private
  communication. See also
  \citet{2014MNRAS.440..663Z,2014ApJ...787...18A}. and will be probed
  in the next few years by Fermi-LAT. } The spectral index of the CRp
distribution is relatively steep ($\alpha_{\rmn{p}}\sim2.6$) for the
CRp energies $E \gtrsim 10$~GeV that are relevant for the injection of
radio-emitting secondary CRes. The steep spectrum is ultimately a
consequence of the shock history of the simulated cluster, with a weak
dependence on our test particle model for Fermi-I acceleration
\citep{pinzke13}, where we steepen the spectral index to avoid
acceleration efficiencies above $\zeta_{\rmn{p}} \sim 10\%$.
